---
title: "Demonstration of rSOILWAT2 (User Manual)"
author: "Daniel R Schlaepfer"
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    toc: true
    toc_depth: 2
#  rmarkdown::html_vignette:
#    toc: true
#    toc_depth: 3
#    #toc_float: true
#    collapsed: false
#    smooth_scroll: false
#    #theme: united
#    #highlight: tango
#    smart: false
vignette: >
  %\VignetteIndexEntry{rSOILWAT2_demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
\pagebreak


# General setup, installation, and R recap

## Install a sufficiently recent rSOILWAT2 package version:
```{r, install, results = "hide"}
  v <- try(utils::packageVersion("rSOILWAT2"), silent = TRUE)
  if (inherits(v, "try-error") || v < "3.1.4") {
    # Download rSOILWAT2 package code from github repository
    args <- paste("clone -b feature_fun_from_rSFSW2",
      "--single-branch --recursive",
      "https://github.com/DrylandEcology/rSOILWAT2.git rSOILWAT2")
    system2(command = "git", args = args)
    # Install package and build vignettes
    devtools::install(build_vignettes = TRUE)
  }

  stopifnot(require("rSOILWAT2", quietly = TRUE))
```
  

## How to find help and documentation
```{r, help, results = "hide"}
  # Package documention
  help(package = "rSOILWAT2")
  
  # Documentation of a specific function, e.g.,
  ?sw_exec
  
  # Code of a specifici function, e.g.,
  sw_exec
```

## Quick R re-cap: how to subset data.frames, matrices, and lists
```{r, subset, results = "hide"}
  ?Extract
  
  #--- Subset a list
  x <- list(ID = 1, Lyr_1 = 1:10, Lyr_2 = 11:30)
  
  x[[1]] # extract element by position
  x[["Lyr_1"]] # extract element by name
  
  x[1:2] # subset by position
  
  
  #--- Subset a data.frame
  x <- as.data.frame(x)
  
  # extract column by column number
  x[[1]]
  x[, 1]
  
  # subset by column numbers
  x[1:2]
  x[, 1:2]
  
  # extract column by column name
  x[["Lyr_1"]]
  x[, "Lyr_1"]
  x$Lyr_1

  
  #--- Subset a matrix
  x <- as.matrix(x)
  
  x[, 1] # extract element by position
  x[, "Lyr_1"] # extract element by name
  
  x[, 1:2] # subset by position
```



# Run example rSOILWAT2 simulation
```{r, simulate0}
  # Load data for an example site: see ?sw_exampleData
  sw_in0 <- rSOILWAT2::sw_exampleData
  # Run SOILWAT2 for the example site
  sw_out0 <- sw_exec(inputData = sw_in0)
```

```{r, exploreS4}
  class(sw_out0)
  slotNames(sw_out0)
```


Task: Create daily/monthly/yearly plots of soil moisture, transpiration,
      evaporation, etc.
```{r, plot1}
  x <- slot(slot(sw_out0, "VWCBULK"), "Day")
  
  ids <- grep("Lyr", colnames(x))
  vwc <- x[, ids]
  matplot(vwc,
    ylim = c(0, max(vwc)),
    xlab = "Time",
    ylab = "Volumetric water content (bulk; m3 / m3)",
    type = "l"
  )
  legend("bottom",
    ncol = 4,
    legend = colnames(vwc),
    lty = 1,
    col = seq_along(ids)
  )
  
  
  x <- slot(slot(sw_out0, "TRANSP"), "Month")
  ids <- grep("transp_total", colnames(x))
  matplot(x[1:24, ids], type = "l")
  
  
  x <- slot(slot(sw_out0, "EVAPSOIL"), "Year")
  ids <- grep("Lyr", colnames(x))
  es <- apply(x[, ids], 1, sum)
  plot(x = x[, "Year"], y = es,
    ylim = c(0, max(es)),
    xlab = "Time",
    ylab = "Bare-soil evaporation (cm)",
    type = "l"
  )
```


## Read inputs from local files on disk (including daily weather forcing)
Task: Use function `sw_inputDataFromFiles` to read inputs from location
      `path_demo` (e.g., this could be inputs for your site)

```{r, path}
  path_demo <- system.file("extdata", "example1", package = "rSOILWAT2")
```

```{r, simulate1}
  # Read data from input files
  sw_in1 <- sw_inputDataFromFiles(dir = path_demo)
  
  # Then run simulation
  sw_out <- sw_exec(inputData = sw_in1)
```


## Read weather data separately from files on disk
```{r, weather1}
  wdata <- getWeatherData_folders(
    LookupWeatherFolder = file.path(path_demo, "Input"),
    weatherDirName = "data_weather"
  )
  
  sw_out <- sw_exec(
    inputData = sw_in1,
    weatherList = wdata
  )
```


# Handling missing weather data

## Read weather data with missing values
Task: use function `getWeatherData_folders` to read weather data with
      missing values from folder `data_weather_missing`
```{r, weather2}
  wdata_gaps <- getWeatherData_folders(
    LookupWeatherFolder = file.path(path_demo, "Input"),
    weatherDirName = "data_weather_missing",
  )
  
  # This will fail: missing data
  sw_out <- try(sw_exec(
    inputData = sw_in1,
    weatherList = wdata_gaps
  ))
```


Option 1: estimate weather generator coefficients with
          function `dbW_estimate_WGen_coefs`,
          generate weather with function `dbW_generateWeather`,
          and run SOILWAT2 with filled-in data
```{r, weather3}
  wgen_coeffs <- dbW_estimate_WGen_coefs(weatherData = wdata)
  
  wdata_filled <- dbW_generateWeather(
    weatherData = wdata_gaps,
    years = swYears_StartYear(sw_in1):swYears_EndYear(sw_in1),
    wgen_coeffs = wgen_coeffs,
    seed = 123
  )
  
  sw_out <- sw_exec(
    inputData = sw_in1,
    weatherList = wdata_filled
  )
```

Option 2: estimate weather generator coefficients and
          run SOILWAT2 with activated weather generator
```{r, weather4}
  sw_in <- sw_in1
  swWeather_UseMarkov(sw_in) <- TRUE
  swMarkov_Prob(sw_in) <- wgen_coeffs[["mkv_doy"]]
  swMarkov_Conv(sw_in) <- wgen_coeffs[["mkv_woy"]]
  
  sw_out <- sw_exec(
    inputData = sw_in,
    weatherList = wdata_gaps
  )
```

# Compare weather and climate based on weather inputs
Task: plot a comparison between `wdata` and `wdata_filled` for
      mean monthly precipitation and mean monthly temperature using the output
      of function `calc_SiteClimate`
```{r, compare1}
  clim1 <- calc_SiteClimate(wdata)
  clim2 <- calc_SiteClimate(wdata_filled)
  
  
  plot(clim1[["meanMonthlyPPTcm"]], clim2[["meanMonthlyPPTcm"]])
  abline(0, 1, col = "red")
  
  plot(clim1[["meanMonthlyTempC"]], clim2[["meanMonthlyTempC"]])
  abline(0, 1, col = "red")
```

Task: compare weather data `wdata` with `wdata_filled` using function
      `compare_weather`
```{r, compare2, results = "hide"}
  compare_weather(
    ref_weather = dbW_weatherData_to_dataframe(wdata),
    weather = dbW_weatherData_to_dataframe(wdata_filled),
    N = 1,
    path = tempdir(),
    tag = "test"
  )
```


# Access other weather data sources
## Download weather data from DayMet
```{r, daymetr, results = "hide"}
  if (!requireNamespace("daymetr", 
    versionCheck = list(op = ">=", version = "1.4"))) {
    install.packages("daymetr")
  }
```

```{r, daymetr_data, results = "hide"}
  dm_Laramie <- daymetr::download_daymet(
    site = "testsite_name",
    lat = 41.3167,
    lon = -105.6833,
    start = 1980,
    end = 2018,
    internal = TRUE,
    simplify = FALSE
  )
```

Task: convert DayMet data `dm_Laramie` into a data.frame with the columns
      "Year", "DOY", "Tmax_C", "Tmin_C", "PPT_cm" and create a rSOILWAT2
      weather object list with function `dbW_dataframe_to_weatherData`
```{r, daymetr_data2}
  vars <- c("year", "yday", "tmax..deg.c.", "tmin..deg.c.", "prcp..mm.day.")
  xdf <- dm_Laramie[["data"]][, vars]
  xdf[, "prcp..mm.day."] <- xdf[, "prcp..mm.day."] / 10 # convert mm -> cm
  colnames(xdf) <- c("Year", "DOY", "Tmax_C", "Tmin_C", "PPT_cm")
  
  wdata_dm <- dbW_dataframe_to_weatherData(weatherDF = xdf)
  
  # This will fail: no leap days
  sw_out <- try(sw_exec(
    inputData = sw_in1,
    weatherList = wdata_dm
  ))
```



Task: Add leap days to data.frame and estimate those with weather generator
```{r, daymetr_data3}
  xdf2 <- dbW_convert_to_GregorianYears(weatherData = wdata_dm)

  wdata_dm2 <- dbW_generateWeather(
    weatherData = dbW_dataframe_to_weatherData(weatherDF = xdf2),
    seed = 123
  )
  
  sw_out <- sw_exec(
    inputData = sw_in1,
    weatherList = wdata_dm2
  )
```


# Specifing soils
```{r, soils}
  sw_in <- sw_in1
  
  soil_has <- swSoils_Layers(sw_in)
  soil_new <- data.frame(soil_has[0, ])
  
  soil_def <- data.frame(
    depth = c(5, 10, 20, 50),
    bulkd = 1.3,
    gravel = 0.1,
    evco = c(0.9, 0.1, 0, 0),
    trco_grass = c(0.1, 0.5, 0.3, 0.1),
    trco_shrub = c(0.1, 0.3, 0.2, 0.4),
    trco_tree = c(0.2, 0.3, 0.3, 0.2),
    trco_forb = c(0.1, 0.3, 0.3, 0.3),
    sand = 0.65,
    clay = 0.05,
    impermbeability = 0,
    soil_temp = 0
  )
  
  soil_new[1:4, ] <- soil_def
  
  swSoils_Layers(sw_in) <- data.matrix(soil_new)
  
  trre <- data.frame(
    ndx = 1:2,
    layer = c(2, 4)
  )
  swSite_TranspirationRegions(sw_in) <- data.matrix(trre)
  
  sw_out <- sw_exec(
    inputData = sw_in,
    weatherList = wdata
  )
```


# Estimating vegetation based on climate relationships

## Vegetation composition (land cover)
Task: estimate vegetation composition with
     function `estimate_PotNatVeg_composition`
```{r, vegcomp}
  clim <- calc_SiteClimate(
    weatherList = wdata,
    do_C4vars = TRUE
  )
  
  
  pnv_comp <- estimate_PotNatVeg_composition(
    MAP_mm = 10 * clim[["MAP_cm"]],
    MAT_C = clim[["MAT_C"]],
    mean_monthly_ppt_mm = 10 * clim[["meanMonthlyPPTcm"]],
    mean_monthly_Temp_C = clim[["meanMonthlyTempC"]],
    dailyC4vars = clim[["dailyC4vars"]]
  )
  
  sw_in <- sw_in1
  ids <- sapply(names(swProd_Composition(sw_in)), function(x) {
    grep(
      pattern = substr(x, 1, 4),
      x = names(pnv_comp[["Rel_Abundance_L1"]]),
      ignore.case = TRUE)
  })
  swProd_Composition(sw_in) <- pnv_comp[["Rel_Abundance_L1"]][ids]
  
  
  sw_out <- sw_exec(
    inputData = sw_in,
    weatherList = wdata
  )
```


# Modifing climate conditions
```{r, climmod}
  sw_in <- sw_in1
  
  # Task: 50% reduction in June-Aug precipitation
  swWeather_MonScalingParams(sw_in)[6:8, "PPT"] <- 0.5
  
  # Task: 2 C warming
  swWeather_MonScalingParams(sw_in)[, c("MaxT", "MinT")] <- 2
  
  
  sw_out <- sw_exec(
    inputData = sw_in,
    weatherList = wdata
  )
```
